The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-28 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Filled in all the text

To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying field of views. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are detected by the \textsc{sumthreshold} method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

@@

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see TODO fix reference Ord et al. (2014). Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format. The output is then directly readable by the most common astronomical software packages.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR (TODO reference?) to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations. For this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; TODO cite), which allows high time resolution observations, might help to analyze these phenomena.

@@

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. A high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, is used to time \textsc{cotter}. The wall-clock runtime for processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel.

@@

\section{Conclusions \& Discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate handling.

\begin{itemize}
 \item Frequency behaviour of instrument makes things harder! Need to fix pass-band correction.
 \item High time resolution
 \item Location of SKA is not free of RFI --> Need good ADCs etc
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
 \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
\end{itemize}

>>>>

To detect RFI in MWA observations, we have used the \textsc{aoflagger} and implemented this as a standard MWA tool to flag all MWA data. The \textsc{aoflagger} allows customizing its flagging strategy, such as changing the threshold levels and expected smoothness of good data. Optimal values vary from telescope to telescope because of different band-pass shapes, different time and frequency resolutions and varying field of views. Specific strategies for several telescopes have been implemented in the \textsc{aoflagger} software, including an MWA specific strategy which is used in this work. The different \textsc{aoflagger} strategies are variations on the original LOFAR strategy described in \citet{lofar-radio-environment}. In the LOFAR strategy, the sky contribution is estimated by applying a 2-dimensional high-pass filter on the visibility amplitudes of each baseline in the time and frequency direction. Subsequently, line-shaped features are detected by the SumThreshold method, which is a combinatorial threshold method \citep{post-correlation-rfi-classification}. After iterating these steps a few times, the scale-invariant rank (SIR) operator is applied on the two-dimensional flag mask. The SIR operator is a morphological technique to search for contaminated samples \citep{scale-invariant-rank-operator}.

@@

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see \citet{ord-2014-mwa-fpga-gpu}. Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format. The output is then directly readable by the most common astronomical software packages.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The SumThreshold and SIR operator algorithms that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR \citep{lofar-imaging-cookbook} to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations. For this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

@@

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; \citealt{mwa-voltage-capture-2014}), which allows high time resolution observations, might help to analyze these phenomena.

@@

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. To time \textsc{cotter}, we use a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup. The wall-clock runtime for processing a single 2~min snapshot of 50~GB with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel.

@@

\section{Conclusions \& Discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate mitigation.

SKA-low will be build at the same location as the MWA, and hence several lessons can be learned from the MWA. First of all, it is clear that the SKA will need to be able to handle some amount of RFI. This includes the availability of computational power to perform the RFI detection, analogue-ditigal converters (ADCs) with sufficient bits and a signal path with headroom in order not to be saturated by RFI TODO say something about base design?. \textsc{cotter} relies on the fact that a standard desktop can still hold a reasonably amount of MWA data in memory for the RFI detection, but because of SKA's large number of elements and high time and frequency resolution, this will be a more challenging problem for the SKA. Flagging MWA data is slightly complicated due to the first poly-phase filter and digital gains. For accurate RFI detection it is best to have a smooth response over a large instantaneous bandwidth.

The \textsc{aoflagger} RFI detection strategy, originally developed for LOFAR, works well for the MWA. Faint RFI events such as the ones in Figs.~\ref{fig:broadband-pulse} and \ref{fig:156_7}, or complex events such as the one in Fig.~\ref{fig:dvb-burst}, are not adequately detected by a single-sample thresholding algorithm, but \textsc{aoflagger}'s SumThreshold and SIR-operator algorithms are able to flag such events. These algorithms have gained some popularity (e.g. TODO Miriad, eMERLIN), but other projects still use single-sample thresholding, e.g. PAPER. Since strong RFI is practically seen everywhere, and because the apparent strength of RFI events will follow a power-law distribution, faint transmitters will interfere with observations of any (terrestrial) radio observatory. Using algorithms with low sensitivity will detect fewer of these, and thus result in RFI becoming more quickly a problem in long-integration projects.

The strategy does not perform well on continuous broadband RFI such as DVB, which is occassionally present due to tropospheric ducting or ionospheric activity. To remove DVB, a second detection round is required, and the current methodology to handle DVB RFI is to delete all affected snapshots, or even the entire night. This requires hardly any computational power, because the visibility statistics are collected in \textsc{cotter}. The current sensitivity of the flagger is enough for calibration and imaging of MWA data. For deep-integration projects, such as the EoR projects, it might be that low-level RFI will show up at lower noise levels. One way to increase the sensitivity of the flagger, is by executing the flagger on the summed power of multiple baselines. Fig.~\ref{fig:156_7} shows that this increases the detectability of an RFI event significantly.
